---
title: |
  | \vspace{7cm} \LARGE ST502: Final Project
author: "Apostolos Stamenos & Tyler Pollard"
date: "4/19/2022"
header-includes:
  - \usepackage{float}
  - \usepackage{indentfirst}
  - \usepackage{caption}
  - \floatplacement{figure}{H}
geometry: "left=1.5cm,right=1.5cm,top=1.5cm,bottom=1.5cm"
linestretch: 1.3
indent: true
output:  pdf_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```
\newpage

\section{Part 2}\label{s:part2}
*Discuss the purpose of simulations (Apostolos)*
*You are expected to discuss why a numerical simulation study can be used to evaluate the two testing procedures*

*(Tyler)Discuss design of (generated data sets) and mentioning how many combinations of the population characteristics with values considered which equals 135 unique data sets eqach containing 100 randomly generated samples for two samples*

\begin{table}[H]\begin{center}
\caption{{\bf Table of blahblahblahblah}: For most of the scenarios, the power of the xxx test is higher than that for the xxx test.}\label{t:power}
\begin{tabular}{c|cc} \hline
Population Characteristics & Values \\ \hline
True Variance & $\sigma_1^2 = 1,4,9$ and $\sigma_2^2 = 1$ \\
Sample Size & $n_1 = 10,30,70$ and $n_2 = 10,30,70$ \\
True Mean Difference & $\Delta = \mu_1-\mu_2 = -5,-1,0,1,5$\\ \hline
\end{tabular}
\end{center}
\end{table}

\captionof{figure}{\bf Caption}
\label{f: alpha}
\centering
```{r alpha, echo=FALSE, warning=FALSE, message = FALSE, fig.width = 5.5, fig.height = 4}
source("hypothesis tests.R")
n1.labs <- c('n1 = 10', 'n1 = 30', 'n1 = 70')
n2.labs <- c('n2 = 10', 'n2 = 30', 'n2 = 70')
names(n1.labs) <- c(10, 30, 70)
names(n2.labs) <- c(10, 30, 70)

# For alpha create 3x3 figure of varying n's and on x axis plot var1 and y axis alpha and parse by test
# Create using points with line
ggplot(data = combined_alpha_df) +
  geom_point(aes(x = var1, y = probs, color = type), size = 1.5) +
  geom_line(aes(x = var1, y = probs, color = type), size = 1) +
  facet_grid(n2 ~ n1, labeller = labeller(
    n1 = n1.labs,
    n2 = n2.labs
  )) +
  scale_x_continuous(breaks = c(1,4,9)) +
  scale_colour_manual(name = "Test", values = c('Pooled' = 'darkorange1', 'Satterthwaite' = 'mediumpurple1')) +
  labs(x = expression(sigma[1]^2), y = expression(paste("Alpha, ", alpha))) +
  theme_bw()
```

\captionof{figure}{\bf Caption}
\label{f: power}
\centering
```{r power, echo=FALSE, warning=FALSE, message = FALSE, fig.width = 5.5, fig.height = 4}
source("hypothesis tests.R")
n1.labs <- c('n1 = 10', 'n1 = 30', 'n1 = 70')
n2.labs <- c('n2 = 10', 'n2 = 30', 'n2 = 70')
names(n1.labs) <- c(10, 30, 70)
names(n2.labs) <- c(10, 30, 70)

# For power create 3x3 figure of varying n's and on x axis plot true mean diff and y axis power and parse by test and var Create 6 line plots for each of the 9 graphs. Different line types for each test and different color for each variance.
ggplot(data = combined_power_df) +
  geom_line(aes(x = delta, y = probs, color = var1, linetype = type), size = 1) +
  facet_grid(n2 ~ n1, labeller = labeller(
    n1 = n1.labs,
    n2 = n2.labs
  )) +
  scale_x_continuous(breaks = c(-5,-1,1,5)) +
  scale_linetype_discrete(name = "Test") +
  scale_colour_manual(name = expression(sigma[1]^2), values = c('1' = 'springgreen4', '4' = 'darkred', '9' = 'steelblue')) +
  labs(x = expression(paste("True Mean Difference, ", Delta)), y = expression(paste('Power = ', 1-beta))) +
  theme_bw()
```



*Results of simulations with the plots for both alpha and power (Apostolos focus on power Tyler on alpha)*
<!-- \begin{figure}\caption{{\bf yadayadayada}\label{f:alpha}} -->
<!-- \centering  -->
<!-- \includegraphics[page=1,width=1.05\textwidth]{alpha} -->
<!-- \end{figure} -->

<!-- \begin{figure}\caption{{\bf yadayadayada}\label{f:power}} -->
<!-- \centering  -->
<!-- \includegraphics[page=1,width=1.05\textwidth]{power} -->
<!-- \end{figure} -->

*Discuss conclusions based on plots and other theoretical arguments (Apostolos)*

